---
title: "Deliverable 3"
author: "Michaela Bowen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Data

```{r, message=FALSE, warning=FALSE}

# Load libraries
library(readr)
library(dplyr)

# Load Spotify data from GitHub
spotify_data <- read_csv("https://raw.githubusercontent.com/Mikay711/DPP_Deliv3/refs/heads/main/Spotify_Cleaned.csv")

# Load Hot100 data from GitHub
hot100_data <- read_csv("https://raw.githubusercontent.com/Mikay711/DPP_Deliv3/refs/heads/main/Hot100_2023_Cleaned.csv")


# Preview the data
head(spotify_data)
head(hot100_data)



```

## Rescaling Spotify Data

In this section, we process the Spotify Most Streamed Songs dataset to:

-   Normalize percentage-based attributes (e.g., `danceability`, `valence`, `energy`) to a scale of 0–1 for consistency.

-   Categorize song stream counts into "low," "medium," and "high" based on thresholds.

```{r, message=FALSE, warning=FALSE}
# Rescale percentage attributes in Spotify data to 0–1
spotify_data <- spotify_data %>%
  mutate(
    danceability_scaled = `danceability_%` / 100,
    valence_scaled = `valence_%` / 100,
    energy_scaled = `energy_%` / 100,
    acousticness_scaled = `acousticness_%` / 100,
    instrumentalness_scaled = `instrumentalness_%` / 100,
    liveness_scaled = `liveness_%` / 100,
    speechiness_scaled = `speechiness_%` / 100
  )

# Ensure 'streams' is numeric
spotify_data <- spotify_data %>%
  mutate(streams = as.numeric(streams))

# Categorize 'streams' into "low," "medium," "high"
spotify_data <- spotify_data %>%
  mutate(
    stream_category = case_when(
      streams < 500000000 ~ "low",
      streams < 1000000000 ~ "medium",
      TRUE ~ "high"
    )
  )

# Check the distribution of stream categories
table(spotify_data$stream_category)

# Preview the updated dataset
head(spotify_data)


```

## Rescaling Hot 100 2023 Data

In this section, we process the Billboard Hot100 dataset to:

-   Handle missing values in the `Weeks in Charts` column.

-   Categorize song ranks into "Top 10," "Top 50," and "Top 100."

-   Group the number of weeks on the charts into meaningful ranges ("1-10," "11-20," etc.).

```{r, message=FALSE, warning=FALSE}

hot100_data <- hot100_data %>%
  mutate(`Weeks in Charts` = ifelse(`Weeks in Charts` == "-", NA, `Weeks in Charts`))

hot100_data <- hot100_data %>%
  mutate(`Weeks in Charts` = as.numeric(`Weeks in Charts`))

# Hot100: Categorize 'Rank' into "Top 10", "Top 50", "Top 100"
hot100_data <- hot100_data %>%
  mutate(
    rank_category = case_when(
      Rank <= 10 ~ "Top 10",
      Rank <= 50 ~ "Top 50",
      TRUE ~ "Top 100"
    )
  )

# Hot100: Group 'Weeks in Charts' into ranges
hot100_data <- hot100_data %>%
  mutate(
    weeks_category = case_when(
      as.numeric(`Weeks in Charts`) <= 10 ~ "1-10",
      as.numeric(`Weeks in Charts`) <= 20 ~ "11-20",
      as.numeric(`Weeks in Charts`) <= 30 ~ "21-30",
      TRUE ~ "31+"
    )
  )

head(hot100_data)
```

## Save Processed Data

```{r, message=FALSE, warning=FALSE}
write_csv(spotify_data, "Processed Data/Spotify_Rescaled.csv")
write_csv(hot100_data, "Processed Data/Hot100_Rescaled.csv")


```
